# Pre-processing

```{r, results='hide', message=FALSE, warning=FALSE}
# Joining the two datasets by building_id

library(tidyverse)
library(tidymodels)

train_values <- read_csv("train_values.csv")
train_labels <- read_csv("train_labels.csv")

earthquake <- train_values |>
  left_join(train_labels, by = "building_id")
```

```{r}
# Check what the data looks like, check for missing values

glimpse(earthquake)
summary(earthquake)
colSums(is.na(earthquake))
```

For this project, building_id has no predictive power, and keeping it would only introduce noise into the dataset, so I decided to drop it. Additionally, has_secondary_use is 1 whenever any detailed secondary-use flag is 1, making it redundant if we include all of the detailed indicators. That's why I'm also dropping that one. 

```{r}
# Drop building_id column, as well as the redundant has_secondary_use column
earthquake <- earthquake |>
  select(-building_id, -has_secondary_use)

# Convert categorical variables to factors
earthquake <- earthquake |>
  mutate(
    land_surface_condition = factor(land_surface_condition),
    foundation_type = factor(foundation_type),
    roof_type = factor(roof_type),
    ground_floor_type = factor(ground_floor_type),
    other_floor_type = factor(other_floor_type),
    position = factor(position),
    plan_configuration = factor(plan_configuration),
    legal_ownership_status = factor(legal_ownership_status),
    damage_grade = factor(damage_grade, levels = c(1, 2, 3), ordered = TRUE),
    geo_level_1_id = factor(geo_level_1_id),
    geo_level_2_id = factor(geo_level_2_id),
    geo_level_3_id = factor(geo_level_3_id)
  )
```

# Exploratory Data Analysis

## Inspecting the numeric variables

```{r}
library(ggplot2)
library(patchwork)
# Inspect distribution

earthquake |>
  select(
    count_floors_pre_eq,
    age,
    area_percentage,
    height_percentage,
    count_families
  ) |>
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "value"
  ) |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ variable, scales = "free")
```

Most distributions are slightly right-skewed, except for age which is extremely right-skewed, with a noticeable number of observations near the 1000 mark, which requires further investigation.

```{r}
# Investigate age
earthquake |>
  count(age) |>
  arrange(desc(age)) |>
  head(20)
```

There's an exceptionally large number of observations at age == 995, which suggests it has a special meaning rather than being the actual age of a building.

```{r}
library(GGally)

# 1. Select numerical predictors with log transformations and damage_grade as numeric
numerical_vars <- earthquake |>
  mutate(
    damage_grade = as.numeric(as.character(damage_grade)),
    log_age = log(age + 1),
    log_height_percentage = log(height_percentage)
  ) |>
  select(
    damage_grade,
    count_floors_pre_eq,
    log_age,
    area_percentage,
    log_height_percentage,
    count_families
  )

# 2. Compute correlation matrix & find the most correlated numerical predictor with damage_grade
correlations_table <- cor(numerical_vars, use = "pairwise.complete.obs")
correlations <- correlations_table["damage_grade", ]
correlations <- correlations[names(correlations) != "damage_grade"]

# Predictor with strongest correlation:
most_correlated <- names(correlations)[which.max(abs(correlations))]
most_correlated

# 3. Pairwise plot with damage_grade and numerical predictors
# (Sampling rows is recommended if dataset is large, e.g. slice_sample(n = 2000))
GGally::ggpairs(
  numerical_vars |> slice_sample(n = min(2000, nrow(numerical_vars))),
  lower = list(
    continuous = GGally::wrap(
      "smooth", method = "lm", se = FALSE, alpha = 0.25
    )
  )
)
```

## Dealing with geo_level_id

While the categorical variable geo_level_1_id only has 31 unique values, geo_level_2_id has over 1400, and geo_level_3_id has over 12,000. My assumption is that geo_level_1_id represents the the largest region, and that observations with the same geo_level_1_id would have different geo_level_2_id, implying a nested nested structure of geographic hierarchy. To test this assumption, I'll check whether there are any observation with the same geo_level_1_id, but different geo_level_2_id, and the same for 2 and 3, as well as 1 and 3

```{r}
# check whether there are any observation with the same geo_level_1_id, but different geo_level_2_id
earthquake |>
  distinct(geo_level_1_id, geo_level_2_id) |>
  count(geo_level_2_id) |>
  filter(n > 1)
```

```{r}
# check whether there are any observation with the same geo_level_2_id, but different geo_level_3_id
earthquake |>
  distinct(geo_level_2_id, geo_level_3_id) |>
  count(geo_level_3_id) |>
  filter(n > 1)
```

```{r}
# check whether there are any observation with the same geo_level_1_id, but different geo_level_3_id
earthquake |>
  distinct(geo_level_1_id, geo_level_3_id) |>
  count(geo_level_3_id) |>
  filter(n > 1)
```

So my assumption was correct, and the geographic ids are nested, now I want to test whether there are substantial variance in levels of damage across geographic regions. We start from large to small (level_1_ids, and then different level_2_ids within the same level_1_ids, and then different level_3_ids within the same level_2_ids)

```{r}
# Investigate whether damage grade differs substantially by geographic region

geo_order <- earthquake |>
  mutate(damage_num = as.numeric(as.character(damage_grade))) |>
  group_by(geo_level_1_id) |>
  summarise(
    mean_damage = mean(damage_num)
  ) |>
  arrange(mean_damage)

geo1_damage <- earthquake |>
  mutate(
    geo_level_1_id = factor(
      geo_level_1_id,
      levels = geo_order$geo_level_1_id
    )
  ) |>
  count(geo_level_1_id, damage_grade) |>
  group_by(geo_level_1_id) |>
  mutate(prop = n / sum(n))

ggplot(
  geo1_damage,
  aes(x = geo_level_1_id, y = prop, fill = damage_grade)
) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Geographic Level 1",
    y = "Proportion of Buildings",
    fill = "Damage Grade"
  ) +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```

Based on the visualization, there IS substantial damage grade variation across geographic regions, which justify keeping the first level of geographic ids.

For level 2, I'll examine all the level-2 regions within the same level-1 region id 6:

```{r}
geo_order <- earthquake |>
  mutate(damage_num = as.numeric(as.character(damage_grade))) |>
  group_by(geo_level_2_id) |>
  summarise(
    mean_damage = mean(damage_num)
  ) |>
  arrange(mean_damage)

geo2_damage <- earthquake |>
  mutate(
    geo_level_2_id = factor(
      geo_level_2_id,
      levels = geo_order$geo_level_2_id
    )
  )|>
  count(
    geo_level_1_id,
    geo_level_2_id,
    damage_grade
  ) |>
  group_by(
    geo_level_1_id,
    geo_level_2_id
  ) |>
  mutate(
    prop = n / sum(n)
  ) |>
  ungroup()

geo2_damage |>
  filter(geo_level_1_id == 6) |>
  ggplot(
    aes(
      x = geo_level_2_id,
      y = prop,
      fill = damage_grade
    )
  ) +
  geom_col() +
  scale_y_continuous(
    labels = scales::percent
  ) +
  labs(
    x = "Geographic Level 2",
    y = "Proportion of Buildings",
    fill = "Damage Grade"
  ) +
  theme(
    axis.text.x = element_text(
      angle = 90,
      hjust = 1
    )
  )
```

Based on the visualization, there is substantial variation in levels of damage within the same level-1 region as well, which justifies keeping the second level of geographical id.

For level 3, I'll examine all the level-3 regions within the same level-2 region id 487:

```{r}
geo_order <- earthquake |>
  mutate(damage_num = as.numeric(as.character(damage_grade))) |>
  group_by(geo_level_3_id) |>
  summarise(
    mean_damage = mean(damage_num)
  ) |>
  arrange(mean_damage)

geo3_damage <- earthquake |>
  mutate(
    geo_level_3_id = factor(
      geo_level_3_id,
      levels = geo_order$geo_level_3_id
    )
  )|>
  count(
    geo_level_2_id,
    geo_level_3_id,
    damage_grade
  ) |>
  group_by(
    geo_level_2_id,
    geo_level_3_id
  ) |>
  mutate(
    prop = n / sum(n)
  ) |>
  ungroup()

geo3_damage |>
  filter(geo_level_2_id == 487) |>
  ggplot(
    aes(
      x = geo_level_3_id,
      y = prop,
      fill = damage_grade
    )
  ) +
  geom_col() +
  scale_y_continuous(
    labels = scales::percent
  ) +
  labs(
    x = "Geographic Level 3",
    y = "Proportion of Buildings",
    fill = "Damage Grade"
  ) +
  theme(
    axis.text.x = element_text(
      angle = 90,
      hjust = 1
    )
  )
```

There is variation among the third level of geographic id as well, but seemingly not as pronounced as the first two levels. However, I don't know if every second-level region looks like this, it got me thinking: On average, how many third-level region is in one second-level region? And how many second-level region is in one first-level region? Because here's the rationale: the more sub-regions are contained within one region, the more likely the effect of an earthquake will be varied across said sub-regions. 

That's why I'll count how many level-2 regions is inside each level-1 region, and how many level-3 region is inside each level-2 region

```{r}
# counting level-2 regions
geo1_counts <- earthquake |>
  group_by(geo_level_1_id) |>
  summarise(
    n_geo_level_2 = n_distinct(geo_level_2_id),
    .groups = "drop"
  ) |>
  arrange(desc(n_geo_level_2))

geo1_counts
summary(geo1_counts$n_geo_level_2)
```

```{r}
# counting level-3 regions

geo2_counts <- earthquake |>
  group_by(
    geo_level_1_id,
    geo_level_2_id
  ) |>
  summarise(
    n_geo_level_3 = n_distinct(geo_level_3_id),
    .groups = "drop"
  ) |>
  arrange(desc(n_geo_level_3))

geo2_counts
summary(geo2_counts$n_geo_level_3)
```

So on average, each level-1 region contains around 46 level-2 regions, while each level-2 regions only contain 8 level-3 regions. For this reason, I think there is definitely an argument to roll the level-3 geographical data into fewer categories, or just outright dropping it.



```{r}
# Order geo_level_1_id by overall mean damage
geo1_order <- earthquake |>
  mutate(damage_num = as.numeric(as.character(damage_grade))) |>
  group_by(geo_level_1_id) |>
  summarise(mean_damage_geo1 = mean(damage_num), .groups = "drop") |>
  arrange(mean_damage_geo1)

# Order geo_level_2_id hierarchically:
# First by geo_level_1_id mean damage, then by geo_level_2_id mean damage within each level 1
geo2_order <- earthquake |>
  mutate(damage_num = as.numeric(as.character(damage_grade))) |>
  group_by(geo_level_1_id, geo_level_2_id) |>
  summarise(mean_damage_geo2 = mean(damage_num), .groups = "drop") |>
  left_join(geo1_order, by = "geo_level_1_id") |>
  arrange(mean_damage_geo1, mean_damage_geo2)

# Compute proportions of damage grade per geo_level_2_id
geo2_damage_all <- earthquake |>
  mutate(
    geo_level_1_id = factor(geo_level_1_id, levels = geo1_order$geo_level_1_id),
    geo_level_2_id = factor(geo_level_2_id, levels = geo2_order$geo_level_2_id)
  ) |>
  count(geo_level_1_id, geo_level_2_id, damage_grade) |>
  group_by(geo_level_1_id, geo_level_2_id) |>
  mutate(prop = n / sum(n)) |>
  ungroup()

# Plot faceted across level 1 regions (preserving level 1 order)
ggplot(
  geo2_damage_all,
  aes(x = geo_level_2_id, y = prop, fill = damage_grade)
) +
  geom_col(position = "stack") +
  facet_wrap(~ geo_level_1_id, scales = "free_x") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Geographic Level 2 (grouped by Geographic Level 1)",
    y = "Proportion of Buildings",
    fill = "Damage Grade"
  ) +
  theme(
    axis.text.x = element_blank(),  # hides tick labels for readability due to 1400+ IDs
    axis.ticks.x = element_blank()
  )
```

```{r}
# Level 3 regions within Level 2 groups, filtered for Level 1 group 2
# We can use another grou, this one just seemed to follow the overall trend of the data

# Filter data for geo_level_1_id == 2
earthquake_geo1_2 <- earthquake |>
  filter(geo_level_1_id == 2)

#  Order geo_level_2_id within geo_level_1_id == 2 by mean damage
geo2_order_g2 <- earthquake_geo1_2 |>
  mutate(damage_num = as.numeric(as.character(damage_grade))) |>
  group_by(geo_level_2_id) |>
  summarise(mean_damage_geo2 = mean(damage_num), .groups = "drop") |>
  arrange(mean_damage_geo2)

#  Order geo_level_3_id hierarchically:
#    First by geo_level_2_id mean damage, then by geo_level_3_id mean damage within each level 2
geo3_order_g2 <- earthquake_geo1_2 |>
  mutate(damage_num = as.numeric(as.character(damage_grade))) |>
  group_by(geo_level_2_id, geo_level_3_id) |>
  summarise(mean_damage_geo3 = mean(damage_num), .groups = "drop") |>
  left_join(geo2_order_g2, by = "geo_level_2_id") |>
  arrange(mean_damage_geo2, mean_damage_geo3)

# Compute proportions of damage grade per geo_level_3_id
geo3_damage_g2 <- earthquake_geo1_2 |>
  mutate(
    geo_level_2_id = factor(geo_level_2_id, levels = geo2_order_g2$geo_level_2_id),
    geo_level_3_id = factor(geo_level_3_id, levels = geo3_order_g2$geo_level_3_id)
  ) |>
  count(geo_level_2_id, geo_level_3_id, damage_grade) |>
  group_by(geo_level_2_id, geo_level_3_id) |>
  mutate(prop = n / sum(n)) |>
  ungroup()

# Plot faceted across level 2 regions within level 1 group 2
ggplot(
  geo3_damage_g2,
  aes(x = geo_level_3_id, y = prop, fill = damage_grade)
) +
  geom_col(position = "stack") +
  facet_wrap(~ geo_level_2_id, scales = "free_x") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Level 3 Damage Proportions within Level 2 Regions (for Level 1 = 2)",
    x = "Geographic Level 3 (grouped by Geographic Level 2)",
    y = "Proportion of Buildings",
    fill = "Damage Grade"
  ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```

```{r}
# Summary statistics on building counts (rows) per geo_level_3_id
geo3_row_counts <- earthquake |>
  count(geo_level_3_id, name = "n_buildings")

summary(geo3_row_counts$n_buildings)

# Distribution plot of building counts per level 3 region
ggplot(geo3_row_counts, aes(x = n_buildings)) +
  geom_histogram(bins = 50) +
  scale_x_log10() +
  labs(
    title = "Distribution of Building Counts per Geo Level 3 Region (log10 scale)",
    x = "Number of Buildings (Rows)",
    y = "Count of Level 3 Regions"
  )
```

## Inspecting the remaining categorical variables

```{r, fig.width=14, fig.height=14, out.width="100%"}
# Select categorical variables (excluding high-cardinality geo_level IDs) and numeric damage_grade
cat_vars <- c(
  "land_surface_condition",
  "foundation_type",
  "roof_type",
  "ground_floor_type",
  "other_floor_type",
  "position",
  "plan_configuration",
  "legal_ownership_status"
)

cat_data <- earthquake |>
  mutate(damage_grade_num = as.numeric(as.character(damage_grade))) |>
  select(all_of(cat_vars), damage_grade_num)

# GGally ggpairs with boxplots for categorical vs numeric damage_grade
GGally::ggpairs(
  cat_data |> slice_sample(n = min(2000, nrow(cat_data))),
  columns = c(cat_vars, "damage_grade_num"),
  lower = list(
    combo = GGally::wrap("box_no_facet", alpha = 0.25)
  )
)
```

```{r, fig.width=12, fig.height=10}
# Compute correlation ratio (eta = sqrt(R^2)) for each categorical variable
cat_correlations <- purrr::map_dfr(cat_vars, function(var) {
  model <- lm(as.formula(paste("damage_grade_num ~", var)), data = cat_data)
  r_squared <- summary(model)$r.squared
  tibble::tibble(
    variable = var,
    eta = sqrt(r_squared),
    r_squared = r_squared
  )
}) |>
  arrange(desc(eta))

# Compute category-level summary stats (mean damage & frequency)
category_level_stats <- purrr::map_dfr(cat_vars, function(var) {
  earthquake |>
    mutate(damage_num = as.numeric(as.character(damage_grade))) |>
    group_by(category_level = as.character(.data[[var]])) |>
    summarise(
      variable = var,
      n = n(),
      mean_damage = mean(damage_num),
      prop_grade_3 = mean(damage_num == 3),
      .groups = "drop"
    )
}) |>
  arrange(desc(mean_damage))

# Join overall eta to category-level stats
plot_cat_data <- category_level_stats |>
  left_join(cat_correlations, by = "variable") |>
  mutate(
    facet_label = paste0(variable, " (eta = ", round(eta, 3), ")"),
    facet_label = reorder(facet_label, -eta)
  )

# Plot categories ranked by mean damage with eta annotated
ggplot(plot_cat_data, aes(x = reorder(category_level, mean_damage), y = mean_damage, fill = mean_damage)) +
  geom_col() +
  geom_text(
    aes(label = sprintf("%.2f (n=%d)", mean_damage, n)),
    hjust = 1.1,
    color = "white",
    size = 3
  ) +
  coord_flip() +
  facet_wrap(~ facet_label, scales = "free_y") +
  scale_fill_viridis_c(option = "plasma", name = "Mean Damage") +
  labs(
    title = "Categorical Predictors Ranked by Association with Damage Grade",
    subtitle = "Faceted by variable (ordered by eta); categories sorted by mean damage grade (1 to 3)",
    x = "Category Level",
    y = "Mean Damage Grade (1 = Low, 3 = High)"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))
```

# Train-test split

Because Grade 2 is the most common damage level and grade 1 the least common, I'll split the data 80/20 into the training and test set, stratified on damage_grade

```{r}

set.seed(98)

earthquake_split <- initial_split(earthquake, prop = 0.8, strata = damage_grade)

train <- training(earthquake_split)
test <- testing(earthquake_split)
```


