# Pre-processing

```{r, results='hide', message=FALSE, warning=FALSE}
# Joining the two datasets by building_id

library(tidyverse)
library(tidymodels)

train_values <- read_csv("train_values.csv")
train_labels <- read_csv("train_labels.csv")

earthquake <- train_values |>
  left_join(train_labels, by = "building_id")
```

```{r}
# Check what the data looks like, check for missing values

glimpse(earthquake)
summary(earthquake)
colSums(is.na(earthquake))
```

For this project, building_id has no predictive power, and keeping it would only introduce noise into the dataset, so I decided to drop it. Additionally, has_secondary_use is 1 whenever any detailed secondary-use flag is 1, making it redundant if we include all of the detailed indicators. That's why I'm also dropping that one. 

```{r}
# Drop building_id column, as well as the redundant has_secondary_use column
earthquake <- earthquake |>
  select(-building_id, -has_secondary_use)

# Convert categorical variables to factors
earthquake <- earthquake |>
  mutate(
    land_surface_condition = factor(land_surface_condition),
    foundation_type = factor(foundation_type),
    roof_type = factor(roof_type),
    ground_floor_type = factor(ground_floor_type),
    other_floor_type = factor(other_floor_type),
    position = factor(position),
    plan_configuration = factor(plan_configuration),
    legal_ownership_status = factor(legal_ownership_status),
    damage_grade = factor(damage_grade, levels = c(1, 2, 3), ordered = TRUE),
    geo_level_1_id = factor(geo_level_1_id),
    geo_level_2_id = factor(geo_level_2_id),
    geo_level_3_id = factor(geo_level_3_id)
  )
```

# Exploratory Data Analysis

## Inspecting the numeric variables

```{r}
library(ggplot2)
library(patchwork)
# Inspect distribution

earthquake |>
  select(
    count_floors_pre_eq,
    age,
    area_percentage,
    height_percentage,
    count_families
  ) |>
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "value"
  ) |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ variable, scales = "free")
```

Most distributions are slightly right-skewed, except for age which is extremely right-skewed, with a noticeable number of observations near the 1000 mark, which requires further investigation.

```{r}
# Investigate age
earthquake |>
  count(age) |>
  arrange(desc(age)) |>
  head(20)
```

There's an exceptionally large number of observations at age == 995, which suggests it has a special meaning rather than being the actual age of a building.

## Dealing with geo_level_id

While the categorical variable geo_level_1_id only has 31 unique values, geo_level_2_id has over 1400, and geo_level_3_id has over 12,000. My assumption is that geo_level_1_id represents the the largest region, and that observations with the same geo_level_1_id would have different geo_level_2_id, implying a nested nested structure of geographic hierarchy. To test this assumption, I'll check whether there are any observation with the same geo_level_1_id, but different geo_level_2_id, and the same for 2 and 3, as well as 1 and 3

```{r}
# check whether there are any observation with the same geo_level_1_id, but different geo_level_2_id
earthquake |>
  distinct(geo_level_1_id, geo_level_2_id) |>
  count(geo_level_2_id) |>
  filter(n > 1)
```

```{r}
# check whether there are any observation with the same geo_level_2_id, but different geo_level_3_id
earthquake |>
  distinct(geo_level_2_id, geo_level_3_id) |>
  count(geo_level_3_id) |>
  filter(n > 1)
```

```{r}
# check whether there are any observation with the same geo_level_1_id, but different geo_level_3_id
earthquake |>
  distinct(geo_level_1_id, geo_level_3_id) |>
  count(geo_level_3_id) |>
  filter(n > 1)
```

So my assumption was correct, and the geographic ids are nested, now I want to test whether there are substantial variance in levels of damage across geographic regions. We start from large to small (level_1_ids, and then different level_2_ids within the same level_1_ids, and then different level_3_ids within the same level_2_ids)

```{r}
# Investigate whether damage grade differs substantially by geographic region

geo_order <- earthquake |>
  mutate(damage_num = as.numeric(as.character(damage_grade))) |>
  group_by(geo_level_1_id) |>
  summarise(
    mean_damage = mean(damage_num)
  ) |>
  arrange(mean_damage)

geo1_damage <- earthquake |>
  mutate(
    geo_level_1_id = factor(
      geo_level_1_id,
      levels = geo_order$geo_level_1_id
    )
  ) |>
  count(geo_level_1_id, damage_grade) |>
  group_by(geo_level_1_id) |>
  mutate(prop = n / sum(n))

ggplot(
  geo1_damage,
  aes(x = geo_level_1_id, y = prop, fill = damage_grade)
) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Geographic Level 1",
    y = "Proportion of Buildings",
    fill = "Damage Grade"
  ) +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```

Based on the visualization, there IS substantial damage grade variation across geographic regions, which justify keeping the first level of geographic ids.

For level 2, I'll examine all the level-2 regions within the same level-1 region id 6:

```{r}
geo_order <- earthquake |>
  mutate(damage_num = as.numeric(as.character(damage_grade))) |>
  group_by(geo_level_2_id) |>
  summarise(
    mean_damage = mean(damage_num)
  ) |>
  arrange(mean_damage)

geo2_damage <- earthquake |>
  mutate(
    geo_level_2_id = factor(
      geo_level_2_id,
      levels = geo_order$geo_level_2_id
    )
  )|>
  count(
    geo_level_1_id,
    geo_level_2_id,
    damage_grade
  ) |>
  group_by(
    geo_level_1_id,
    geo_level_2_id
  ) |>
  mutate(
    prop = n / sum(n)
  ) |>
  ungroup()

geo2_damage |>
  filter(geo_level_1_id == 6) |>
  ggplot(
    aes(
      x = geo_level_2_id,
      y = prop,
      fill = damage_grade
    )
  ) +
  geom_col() +
  scale_y_continuous(
    labels = scales::percent
  ) +
  labs(
    x = "Geographic Level 2",
    y = "Proportion of Buildings",
    fill = "Damage Grade"
  ) +
  theme(
    axis.text.x = element_text(
      angle = 90,
      hjust = 1
    )
  )
```

Based on the visualization, there is substantial variation in levels of damage within the same level-1 region as well, which justifies keeping the second level of geographical id.

For level 3, I'll examine all the level-3 regions within the same level-2 region id 487:

```{r}
geo_order <- earthquake |>
  mutate(damage_num = as.numeric(as.character(damage_grade))) |>
  group_by(geo_level_3_id) |>
  summarise(
    mean_damage = mean(damage_num)
  ) |>
  arrange(mean_damage)

geo3_damage <- earthquake |>
  mutate(
    geo_level_3_id = factor(
      geo_level_3_id,
      levels = geo_order$geo_level_3_id
    )
  )|>
  count(
    geo_level_2_id,
    geo_level_3_id,
    damage_grade
  ) |>
  group_by(
    geo_level_2_id,
    geo_level_3_id
  ) |>
  mutate(
    prop = n / sum(n)
  ) |>
  ungroup()

geo3_damage |>
  filter(geo_level_2_id == 487) |>
  ggplot(
    aes(
      x = geo_level_3_id,
      y = prop,
      fill = damage_grade
    )
  ) +
  geom_col() +
  scale_y_continuous(
    labels = scales::percent
  ) +
  labs(
    x = "Geographic Level 3",
    y = "Proportion of Buildings",
    fill = "Damage Grade"
  ) +
  theme(
    axis.text.x = element_text(
      angle = 90,
      hjust = 1
    )
  )
```

There is variation among the third level of geographic id as well, but seemingly not as pronounced as the first two levels. However, I don't know if every second-level region looks like this, it got me thinking: On average, how many third-level region is in one second-level region? And how many second-level region is in one first-level region? Because here's the rationale: the more sub-regions are contained within one region, the more likely the effect of an earthquake will be varied across said sub-regions. 

That's why I'll count how many level-2 regions is inside each level-1 region, and how many level-3 region is inside each level-2 region

```{r}
# counting level-2 regions
geo1_counts <- earthquake |>
  group_by(geo_level_1_id) |>
  summarise(
    n_geo_level_2 = n_distinct(geo_level_2_id),
    .groups = "drop"
  ) |>
  arrange(desc(n_geo_level_2))

geo1_counts
summary(geo1_counts$n_geo_level_2)
```

```{r}
# counting level-3 regions

geo2_counts <- earthquake |>
  group_by(
    geo_level_1_id,
    geo_level_2_id
  ) |>
  summarise(
    n_geo_level_3 = n_distinct(geo_level_3_id),
    .groups = "drop"
  ) |>
  arrange(desc(n_geo_level_3))

geo2_counts
summary(geo2_counts$n_geo_level_3)
```

So on average, each level-1 region contains around 46 level-2 regions, while each level-2 regions only contain 8 level-3 regions. For this reason, I think there is definitely an argument to roll the level-3 geographical data into fewer categories, or just outright dropping it.

## Inspecting the remaining categorical variables (to be added later)

# Train-test split

Because Grade 2 is the most common damage level and grade 1 the least common, I'll split the data 80/20 into the training and test set, stratified on damage_grade

```{r}

set.seed(98)

earthquake_split <- initial_split(earthquake, prop = 0.8, strata = damage_grade)

train <- training(earthquake_split)
test <- testing(earthquake_split)
```


